---
title: "Reproducible Research Course Project 1: Peer Graded Assessment"
author: "U.Esparza"
date: "2020/06/14"
output:
  html_document: default
  pdf_document: default
---

---

# 1) Loading and pre-processing the data

### Firstly, the data must be read:

##### If not done yet, the zip file must be unzipped first. To do so, the below unzip() function can be used by removing the # symbol from the beginning of the code-line:

```{r, echo=TRUE}
# unzip(zipfile = "./data/repdata_data_activity.zip", exdir = "./data")
reading <- read.csv("./data/activity.csv", sep=",")
```

### Secondly, the dataset is checked using str(), head() and tail() functions:

```{r, echo=TRUE}
str(reading)
head(reading)
tail(reading)
```

### Thirdly, the dates currently in character format must be transformed into POSIXct format by using strtime() and as.POSIXct():

```{r, echo=TRUE}
reading$date <- as.POSIXct(strptime(reading$date, "%Y-%m-%d"))
```


---
 
# 2) Histogram of the total number of steps taken each day

### Firstly, the ggplot2 and dplyr packages must be loaded:
##### If not done yet, the ggplot2 and dplyr packages must be installed. This can be done by removing the # symbol from the beginning of the function install.packages() in the fisrt two code-lines below:

```{r, echo=TRUE}
# install.packages("ggplot2")
# install.packages("dplyr")
library(ggplot2)
library(dplyr)
```

### Secondly, a new variable summarizing the total steps per day must be created:

```{r, echo=TRUE}
mySumPerDate <- reading %>% group_by(date) %>% summarize(sumSteps=sum(steps))
```

### Thirdly, the histogram with the total number of steps per day is plotted:

```{r, echo=TRUE}
ggplot(mySumPerDate, aes(x = date, y = sumSteps), stat = "bin") + 
geom_histogram(stat = "identity", col = "blue", bg = "gray") +
ggtitle("Total number of steps per day") + ylab("Number of steps")

```

##### Remark: my language settings are set to Basque Language. That's why I get x-axis labels "Urr" (short for Urria, i.e. October in Basque), "Aza" (short for Azaroa, i.e. November), "Abe" (short for Abendua, i.e. December) etc.



---
 
# 3) Mean and median number of steps taken each day

### The mean() and median() functions are used together with the condition na.rm=TRUE to remove NA values:


```{r, echo=TRUE}
mean(mySumPerDate$sumSteps, na.rm=TRUE)
median(mySumPerDate$sumSteps, na.rm=TRUE)

```



---
 
# 4) Time series plot of the average number of steps taken

### Firstly, a new variable is created summarizing the average number of steps per interval:

```{r, echo=TRUE}
myMeanPerInt <- reading %>%  group_by(interval) %>% summarize(meanSteps=mean(steps,na.rm=TRUE))

```

### Secondly, a time series plot is created using the type = "l" condition:

```{r, echo=TRUE}
plot(x=myMeanPerInt$interval, y=myMeanPerInt$meanSteps, type = "l", 
     main = "Time series of average nr. of steps per interval across all days",
     xlab = "5 min. intervals (the values on the axis represent minutes)", 
     ylab = "Nr. of steps (mean)",
     col = "blue")

```


---

# 5) The 5-minute interval that, on average, contains the maximum number of steps

### the max() function is used to get the interval with the maximum nr. of steps, in combination with the row() function to get the complete row (including the interval itself):

```{r, echo=TRUE}
myMeanPerInt[row(myMeanPerInt)[myMeanPerInt==max(myMeanPerInt$meanSteps)],]
```


---

# 6) Code to describe and show a strategy for imputing missing data

### The dataset contains some missing values. Let's check first how many NAs exist with the negated (!) version of the complete.cases() function:

```{r, echo=TRUE}

sum(!complete.cases(reading))
```


### Data imputation consists on inferring the NAs from the known part of the data, which is often a better strategy than just removing NAs.
### The easiest way to carry out an imputation is by using univariate imputing. For instance, NAs can be imputed with a provided constant value, by using statistic calculations as mean, median etc. or by sampling with replacement from non-missing values.
### The strategy used here is the UNIVARIATE IMPUTING BY SAMPLING WITH REPLACEMENT from the non-missing values.
##### In order to use the imputeUnivariate() function, the "missRanger" package must be installed first.To do so, the below install.packages() function can be used by removing the # symbol from the beginning of the code-line:

```{r, echo=TRUE}

# install.packages("missRanger")
library(missRanger)
readingImputed <- imputeUnivariate(x=reading, v = NULL, seed = NULL)

```

### Now we can use the negated (!) version of the complete.cases() function again to check if the imputation actually worked (the result should be zero):

```{r, echo=TRUE}

sum(!complete.cases(readingImputed))

```


---

# 7) Histogram of the total number of steps taken each day after missing values are imputed

### Firstly, the total number of steps per day is calculated and plotted again as done in step 2, but this time using imputed values:

```{r, echo=TRUE}

readingImputed$date <- as.POSIXct(strptime(readingImputed$date, "%Y-%m-%d"))
mySumPerDate <- readingImputed %>% group_by(date) %>% summarize(sumSteps=sum(steps))

ggplot(mySumPerDate, aes(x = date, y = sumSteps), stat = "bin") + 
geom_histogram(stat = "identity", col = "blue", bg = "gray") +
ggtitle("Total number of steps per day (NAs imputed)") + ylab("Number of steps")

```

### As one can see, the plot is slightly different from the one obtained in step 2: as a consequence of the NA imputation process, this time there is no date with missing data.

### Secondly, the mean and the median are calculated again as done in step 2, but this time using imputed data:

```{r, echo=TRUE}

mean(mySumPerDate$sumSteps, na.rm=TRUE)
median(mySumPerDate$sumSteps, na.rm=TRUE)

```

### As one can see, the results are slightly different from the ones achieved in step 2.


---

# 8) Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends.

### Firstly, a factor must be created to differentiate between weekdays and weekends. To do so, the wday() function will be used.

##### In order to use the wday() function, the "lubridate" package must be installed first. If not done yet, the below install.packages() function can be used by removing the # symbol from the beginning of the code-line:

```{r, echo=TRUE}
# install.packages("lubridate")
library(lubridate)
day <- wday(readingImputed$date)


for(i in 1:length(day))          
if (day[i] == 1 | day[i] == 7) {
  day[i] <- "weekend" 
  } else {
    day[i] <- "weekday"
  }

readingDay <- cbind(readingImputed,day)
readingWD <- readingDay[day == "weekday",]
readingWE <- readingDay[day == "weekend",]
  
myMeanPerIntWD <- readingWD %>%  group_by(interval) %>% summarize(meanStepsWD=mean(steps,na.rm=TRUE))

myMeanPerIntWE <- readingWE %>%  group_by(interval) %>% summarize(meanStepsWE=mean(steps,na.rm=TRUE))

```


### Secondly, the panel plot is created:

```{r, echo=TRUE}

par(mfrow = c(1,2))
plot(x=myMeanPerIntWD$interval, y=myMeanPerIntWD$meanStepsWD, type = "l",
     main = "WEEKDAYS", sub = "AVG Nr. of steps per interval",
     xlab = "5 min. intervals", 
     ylab = "Nr. of steps (mean)",
     col = "blue")
plot(x=myMeanPerIntWE$interval, y=myMeanPerIntWE$meanStepsWE, type = "l",
     main = "WEEKENDS", sub = "AVG Nr. of steps per interval",
     xlab = "5 min. intervals", 
     ylab = "Nr. of steps (mean)",
     col = "blue")
```


---

# 9) All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

### The R code needed to reproduce the results is described in the previous steps (from 1 to 8).

---
